---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# ggridges: Ridgeline plots in ggplot2

<!-- badges: start -->
[![Build Status](https://travis-ci.org/wilkelab/ggridges.svg?branch=master)](https://travis-ci.org/wilkelab/ggridges)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/wilkelab/ggridges?branch=master&svg=true)](https://ci.appveyor.com/project/clauswilke/ggridges)
[![Coverage Status](https://img.shields.io/codecov/c/github/clauswilke/ggridges/master.svg)](https://codecov.io/github/clauswilke/ggridges?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/ggridges)](https://CRAN.R-project.org/package=ggridges)
[![CRAN_Downloads_Badge](http://cranlogs.r-pkg.org/badges/ggridges)](http://cranlogs.r-pkg.org/downloads/total/last-month/ggridges)
<!-- badges: end -->

Ridgeline plots are partially overlapping line plots that create the impression of a mountain range. They can be quite useful for visualizing changes in distributions over time or space. These types of plots have also been called ["joyplots"](
https://twitter.com/JennyBryan/status/856674638981550080), in reference to the [iconic cover art](https://blogs.scientificamerican.com/sa-visual/pop-culture-pulsar-origin-story-of-joy-division-s-unknown-pleasures-album-cover-video/) for Joy Division's album _Unknown Pleasures_. However, given the [unfortunate origin](https://en.wikipedia.org/wiki/House_of_Dolls) of the name Joy Division, the term "joyplot" is now discouraged.

## Installation

Please install the stable release from CRAN:

```{r eval = FALSE}
install.packages("ggridges")
```

Alternatively, you can install the latest development version from github:

```{r eval = FALSE}
remotes::install_github("wilkelab/ggridges")
```

## Usage

```{r diamonds}
library(ggplot2)
library(ggridges)
    
ggplot(diamonds, aes(x = price, y = cut)) +
  geom_density_ridges(scale = 4) + 
  scale_y_discrete(expand = c(0, 0)) +     # will generally have to set the `expand` option
  scale_x_continuous(expand = c(0, 0)) +   # for both axes to remove unneeded padding
  coord_cartesian(clip = "off") + # to avoid clipping of the very top of the top ridgeline
  theme_ridges()
```

## Documentation and Examples

First read the [package vignette.](https://wilkelab.org/ggridges/articles/introduction.html) Then read the [reference manual.](https://wilkelab.org/ggridges/reference/index.html)
