---
title: "Gallery of ggjoy examples"
author: "Claus O. Wilke"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 4.5
    fig_height: 3
vignette: >
  %\VignetteIndexEntry{Gallery of ggjoy examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r echo=FALSE, include=FALSE}
library(ggplot2)
library(ggjoy)
```


## Evolution of movie lengths over time

Data from the IMDB, as provided in the ggplot2movies package.
```{r message=FALSE, warning=FALSE, fig.width = 6, fig.height = 6}
library(ggplot2movies)
ggplot(movies[movies$year>1912,], aes(x = length, y = year, group = year)) +
  geom_joy(scale = 10, size = 0.25, rel_min_height = 0.03) +
  theme_joy() +
  scale_x_continuous(limits=c(1, 200), expand = c(0.01, 0)) +
  scale_y_reverse(breaks=c(2000, 1980, 1960, 1940, 1920, 1900), expand = c(0.01, 0))
```

## Results from Catalan regional elections, 1980-2015

Modified after [a figure](https://twitter.com/marcbeldata/status/888697140268204032) originally created by Marc Belzunces. 
```{r message=FALSE, warning=FALSE, fig.width = 6, fig.height = 8}
library(tidyverse)
library(forcats)
Catalan_elections %>%
  mutate(Alt = (as.numeric(as.factor(Year))-1) %% 2 + 1) %>%
  ggplot(aes(y = as.factor(Year) %>% fct_rev())) +
  geom_joy(aes(x = Percent, fill = paste(Option, Alt)), 
           alpha = .8, color = "white", from = 0, to = 100) +
  labs(x = "Vote (%)",
       y = "Election Year",
       title = "Indy vs Unionist vote in Catalan elections",
       subtitle = "Analysis unit: municipalities (n = 949)",
       caption = "Marc Belzunces (@marcbeldata) | Source: Idescat") +
  scale_y_discrete(expand = c(0.01, 0)) +
  scale_x_continuous(expand = c(0.01, 0)) +
  scale_fill_manual(breaks = c("Indy 1", "Unionist 1"),
                    labels = c(`Indy 1` = "Indy", `Unionist 1` = "Unionist"),
                    values = c("#ff0000", "#ff8080", "#0000ff", "#8080ff"),
                    name = "Option") +
  theme_joy(grid = FALSE)
```

## Temperatures in Lincoln, Nebraska

Modified from a [blog post](http://austinwehrwein.com/data-visualization/it-brings-me-ggjoy/) by Austin Wehrwein.
```{r message=FALSE, fig.width = 7.5, fig.height = 5}
library(viridis)
ggplot(lincoln_weather, aes(x = `Mean Temperature [F]`, y = `Month`, fill = ..x..)) +
  geom_joy_gradient(scale = 3, rel_min_height = 0.01) +
  scale_x_continuous(expand = c(0.01, 0)) +
  scale_y_discrete(expand = c(0.01, 0)) +
  scale_fill_viridis(name = "Temp. [F]", option = "C") +
  labs(title = 'Temperatures in Lincoln NE',
       subtitle = 'Mean temperatures (Fahrenheit) by month for 2016\nData: Original CSV from the Weather Underground') +
  theme_joy(font_size = 13, grid = TRUE) + theme(axis.title.y = element_blank())
```

## Visualization of Poisson random samples with different means

Inspired by a [ggjoy example](https://twitter.com/noamross/status/888405434381545472) by Noam Ross.
```{r message=FALSE, fig.width = 6, fig.height = 6}
set.seed(1234)
pois_data <- data.frame(mean = rep(1:5, each = 10))
pois_data$group <- factor(pois_data$mean, levels=5:1)
pois_data$value <- rpois(nrow(pois_data), pois_data$mean)
ggplot(pois_data, aes(x = value, y = group, group = group)) +
  geom_joy2(stat = "binline", binwidth = 1, scale = 0.75,
          aes(fill = ifelse(as.numeric(group) %% 2 == 0, "light", "dark"))) +
  # dummy text labels just to extend the y axis sufficiently
  geom_text(stat = "bin",
          aes(y = group+0.75*..count../max(..count..)),
          label = "", vjust = 1, nudge_y = 0.2, binwidth = 1) +
  # now the actual text labels
  geom_text(stat = "bin",
          aes(y = group+0.75*..count../max(..count..),
              label = ifelse(..count..>0, ..count.., "")),
          vjust = -.3, size = 4, binwidth = 1) +
  scale_x_continuous(breaks = c(0:12), limits = c(-.5, 13), expand = c(0, 0),
                     name = "random value") +
  scale_y_discrete(expand = c(0.01, 0), name = "Poisson mean",
                   labels = c("5.0", "4.0", "3.0", "2.0", "1.0")) +
  scale_fill_manual(values = c("dark" = "#0000B0", "light" = "#7070D0")) +
  labs(title = "Poisson random samples with different means",
       subtitle = "sample size n=10") +
  guides(y = "none",
         fill = "none") +
  theme_joy(grid = FALSE) +
  theme(axis.title.x = element_text(hjust = 0.5),
        axis.title.y = element_text(hjust = 0.5))
```
