---
title: "Introduction to ggjoy"
author: "Claus O. Wilke"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 5
vignette: >
  %\VignetteIndexEntry{Introduction to ggjoy}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

# Basic usage

The geom `geom_ridgeline` can be used to draw lines with a filled area underneath.
```{r}
library(ggplot2)
library(ggjoy)

data <- data.frame(x = 1:5, y = rep(1, 5), height = c(0, 1, 3, 4, 2))
ggplot(data, aes(x, y, height = height)) + geom_ridgeline()

```


By contrast, the geom `geom_joy` calculates density estimates from the provided data and then plots those, using the ridgeline visualization.
```{r message=FALSE}
ggplot(iris, aes(x=Sepal.Length, y=Species, group=Species)) + geom_joy()
```

There is also `geom_joy2`, which is identical to `geom_joy` except it uses closed polygons instead of ridgelines for drawing.
```{r message=FALSE}
ggplot(iris, aes(x=Sepal.Length, y=Species, group=Species)) + geom_joy2()
```

# Theme

Joyplots tend to require some theme modifications to look good. Most importantly, the y-axis labels should be vertically aligned so that they are flush with the axis ticks rather than vertically centered. The ggjoy package provides a theme `theme_joy` that does this and a few other theme modifications.

```{r message=FALSE}
ggplot(iris, aes(x=Sepal.Length, y=Species, group=Species)) + geom_joy() + theme_joy()
```

However, without any further modifications, there are still a few issues with this plot. First, the ridgeline for the virginica species is slightly cut off at the very top point. Second, the space between the x and y axis labels and the ridgelines is too large. We can fix both issues using the `expand` option for the axis scales.

```{r message=FALSE}
ggplot(iris, aes(x=Sepal.Length, y=Species, group=Species)) + 
  geom_joy() + theme_joy() +
  scale_x_continuous(expand = c(0.01, 0)) +
  scale_y_discrete(expand = c(0.01, 0))
```

By default, `theme_joy` adds a grid, but the grid can be switched off when not needed.
```{r message=FALSE}
ggplot(iris, aes(x=Sepal.Length, y=Species, group=Species)) + 
  geom_joy() + theme_joy(grid = FALSE) +
  scale_x_continuous(expand = c(0.01, 0)) +
  scale_y_discrete(expand = c(0.01, 0))
```

# Gallery of examples

Temperatures in Lincoln, Nebraska. Modified from a [blog post](http://austinwehrwein.com/data-visualization/it-brings-me-ggjoy/) by Austin Wehrwein.
```{r message=FALSE, fig.width = 7.5, fig.height = 5}
ggplot(lincoln_weather, aes(x = `Mean Temperature [F]`, y = `Month`)) +
  geom_joy(scale = 3, rel_min_height = 0.01) +
  scale_x_continuous(expand = c(0.01, 0)) +
  scale_y_discrete(expand = c(0.01, 0)) +
  labs(title = 'Temperatures in Lincoln NE',
       subtitle = 'Mean temperatures (Fahrenheit) by month for 2016\nData: Original CSV from the Weather Underground') +
  theme_joy(font_size = 13, grid = T) + theme(axis.title.y = element_blank())
```

Evolution of movie lengths over time. Data from the IMDB, as provided in the ggplot2movies package.
```{r message=FALSE, fig.width = 6, fig.height = 6}
library(ggplot2movies)
ggplot(movies[movies$year>1912,], aes(x=length, y=year, group=year)) +
  geom_joy(scale=10, size=0.25, rel_min_height=0.03) +
  theme_joy() +
  scale_x_continuous(limits=c(1, 190), expand=c(0.01, 0)) +
  scale_y_reverse(breaks=c(2000, 1980, 1960, 1940, 1920, 1900), expand=c(0.01, 0))
```
